Lowestcase and Uppestcase letters: Advances in Derp Learning

### 2021-03-21 trunk/lowercase/paper/lowercase.tex
Wrote a bunch more...

\usepackage{float}
\usepackage{url}
% \usepackage{code}
% \usepackage{cite}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{chessboard}
% IPA symbols. safe turns off overrides for like \! which I still want
\usepackage[safe]{tipa}

% \usepackage{chessfs}
\usepackage{adjustbox}

@@

\setlength{\footnotesep}{2em}

% \newcommand\comment[1]{}
\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}

@@

\title{Lowestcase and Uppestcase letters: Further adventures in Derp Learning}
\author{Dr.~Tom~Murphy~VII~Ph.D.}\thanks{
Copyright \copyright\ 2021 the Regents of the Wikiplia Foundation.
Appears in SIGBOVIK~2021 with the
OS2TypoLinegap of the Association for Computational Heresy; {\em IEEEEEE!}
press, Verlag-Verlag volume no.~0x40-2A. 1 em}

@@

If so, you made use of the fact that all letters have UPPERCASE
VERSIONS (e.g.~signifying ANGER) and lowercase versions
(e.g.~signifying u dont care lol). These dimensions have other uses,
for example, it is polite to start a person's name with a capital
letter to show that you took the time to regard their humanity (as it
takes extra work to press the caps lock key, press the first letter of
their name, and then press ther caps lock key again to turn it off).
In German, Nouns start with uppercase Letters, signifying Superiority
over other grammatical Categories like Verbs and Adjectives. Lowercase
letters can be used to conserve printer ink. Actually, I'm not sure that
lowercase letters have any other uses, but let's just roll with it.

@@

@@

We want to machine-learn~\cite{neuralnetwork} two functions,
\makelowercase\ and \makeuppercase. Each takes a letterform and
returns a letterform (we can choose how these are represented) and
does the needful, e.g. \makelowercase(\letterform{A}) should return
\letterform{a}. In order to learn this function, we'll at least need a
lot of examples to use as training data. A training example for
\makelowercase\ is a letterform and its expected corresponding
lowercase one. We can ``easily'' find a large amount of examples by
using existing fonts, and pairing their \letterform{A} with their
\letterform{a}, and so on for all 26 letters, and symmetrically for
\makeuppercase.

@@

Next, my lord, do people have bad taste! And I say this as someone who
made dozens of amateurish fonts\cite{dbzfonts} as a high school and
college student and who is contributing several new questionable
fonts\label{sec:newfonts} as a result of this paper. The database is
just filled with garbage that is unusable for this project: Fonts that
are completely illegible, fonts that are missing most of their
characters, fonts with millions of control points, Comic Sans MS,
fonts where every glyph is a drawing of a train, fonts where
everything is fine except that just the lowercase r has a width of
{\tt MAX\_INT}, and so on. So I built a UI
(Figure~\ref{fig:sortition}) for efficiently and mind-numbingly
cleaning up the database by marking fonts as broken or suitable (and
also categorizing them as serif, sans-serif, decorative, techno, etc.,
which I never used). In doing this I noticed another extremely common
problem, which was that many fonts had the same letter shapes for
uppercase and lowercase letters. This would not do for the current
application!

\begin{figure*}[ht]
\includegraphics[width=0.9 \linewidth]{sortition}
\caption{ The interactive font data-cleaning UI. A seemingly endless
  series of fonts presents, with single keypresses putting the fonts
  into common categories such as (b)roken.
  % XXX maybe write more?
} \label{fig:sortition}
\end{figure*}

@@

\paragraph{Black-box optimization.} If you have a function and want
to find arguments that minimize its output, the most efficient
techniques are generally those like gradient descent. (In fact, the
backpropagation algorithm we use to training the neural network in
Section~\ref{sec:neural} is gradient descent on the function that
takes the model weights and produces an error value for each output
node.) The problem with this is that you need to do some math to
compute the gradient function, and anyway you need to deal with fiddly
bits (Section~\ref{sec:fiddly}) unless the function is convex and
smooth, which it will not be. If you don't want to deal with that,
and have a fast computer (and who doesn't?), black-box optimization
algorithms are worth considering. Here, the interface\footnote{
  Here a simplfied wrapper around BiteOpt~\cite{biteopt} in my {\tt cc-lib}
  library. See \url{https://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/cc-lib/opt/}.}
  is just something like (C++):

@@

\begin{figure}[ht]
\includegraphics[width=0.9 \linewidth]{casealignment}
\caption{ Example alignment to reject the font {\tt
    DrippingGooExtended}. At left, \letterform{A} (red) and
  \letterform{a} (green) rendered with the identity transform, and
  their alignment ($35\%$ difference) below. At right, the transform
  found by the black-box optimizer and the resulting alignment with
  $1.7\%$ difference. Note that the shapes are still not an exact
  match (probably errors introduced in the font export process, which
  has to round the data to integers and might apply other non-linear
  transformations like curve simplification), but these are clearly
  not a useful pair for the current
  problem.} \label{fig:casealignment}
\end{figure}

@@

We'll at least put the data in a somewhat normalized form. The neural
network will take a fixed number of inputs to a fixed number of
outputs, so a simple approach is to decide on some maximum number
control points per letter, and only try training on fonts whose
letterforms fit within that budget. Letterforms can be made of
multiple contours (e.g. a stacked \letterform{g} typically has two
holes in it, and \letterform{j} has two disjoint parts). I found that
most clean fonts had three or fewer contours, and when sorting them by
descending length, most fit within 38, 14, and 10 endpoints for the
three. So, I only train on fonts where all of the letters fit within
this budget.\footnote{It would not be a good idea to reject only the letters
that don't fit, because it might result in the network being trained
on more \letterform{l}s (tends to be simple) than \letterform{g}s
(tends to be complex).}

@@

Rather than try to work with both lines and B\'ezier curves, I
normalize each contour to only contain B\'eziers, by turning a line
segment into an equivalent B\'ezier with its control point at the
midpoint. This frees us from having to distinguish the two types in
the data. We also need each of the three contours to not be {\em too
  short}, so I fill out the fixed-size buffers by repeating the last
point. This is not great but does have the correct meaning (bunch of
useless zero-length edges). It has the property that any predicted
data can be rendered and has a straightforward point-by-point error
function (which might not be the case if we were predicting a dynamic
number of points). Finally, I output the contour's points starting
from the point closest to the origin in order to reduce one needless
degree of freedom.

The network I trained has an input layer size of $154 = (38 + 14 + 10)
\times 4 + 3 \times 2$ (one control point and one end point per
B\'ezier curve), plus a starting point for each of the three contours.
The output layer is the same size, plus 26 (see below). There are
three hidden layers of size 308, 308, 360. The first layer is dense
and the remainder are sparse,
%(two thirds to a half of the weights are zero),
for just about 1 million total parameters. All layers are leaky
rectified linear units. If you're taking notes, don't, as again this
does not work well, and I don't know how people figure out what the
right network size is anyway. I just made it up. You can give {\em me}
your notes.

\paragraph{Bonus outputs.} The output includes the predicted shape,
and also 26 individual predictors for each of the 26 letters. So a
training example is actually like \letterform{C} $\rightarrow$
\letterform{c} $[0, 0, 1, 0, 0, \ldots, 0]$, with the 1 in the third
place because C is the third letter. We don't need these outputs for
the problem itself (e.g. to lowercase new letter shapes), but there
are several ideas behind this. First, the lowercasing function we're
trying to learn does take into account the letter of the alphabet
being lowercased. By asking the network to learn this as well (it is
penalized when it gets the prediction wrong), it must learn features
that allow it to distinguish different letters, and those features are
available for use by outputs we {\em do} care about. This is an easy
way to coax it to learn features that I know are meaningful without
having to actually engineer feature extractors by hand (or first train
separate models, etc.). Perhaps the most useful thing is that it's
very clear what the right answer is, so it gives me an easy way to see
if the network is learning anything {\em at all}. (It does.) Finally,
we can do some silly stuff with these; see
Section~\ref{sec:hallucination}.

@@

\begin{figure}[ht]
\includegraphics[width=0.9 \linewidth]{trainingvector}
\caption{ Screenshot of {\tt train.exe} running on the vector-based
  version of the problem. Shown is the \makelowercase\ model's output
  (bottom) on four shapes from four fonts (top). Some dust in between
  is the activation of the network's layers. At the very bottom, the
  26 predictions for ``what letter is this?''. The output for
  \letterform{j} is not too bad; you can see the distinct dot (a
  separate contour) and it sort of looks like a \letterform{j}. The
  \letterform{e} also has two pieces as expected but is otherwise
  garbage. The model is unsure whether the second input is an H
  or a K, and has predicted a shape sort of ambiguously between those
  two. The \lowercase{z} is also an embarrassment.
} \label{fig:trainingvector}
\end{figure}

I was so {\tiny confident} that this wouldn't work that I only trained
a \makelowercase\ model and didn't even worry about the more
complicated simultaneous training setup yet. I ran this for about
22,000 rounds, some 90 million training examples. Indeed it does not
work (Figure~\ref{fig:trainingvector}). It is not a total catastrophe,
though. We can tell from the 26 bonus outputs that the model can
clearly recognize letters (though perhaps just by memorization). Some
of the shapes it generates are along the right lines. ({\em Along the
  right lines}, get it??) I did not feel ANGRY at these results because
I expected it to not really work.

\paragraph{Just try making it more complicated!} This
problem of predicting the vector shape directly is a lot to ask of a
neural network, at least set up this way. One thing that did not sit
well with me is that the network could generate a perfect-looking
result, but because it didn't have the points in the expected order,
it would be penalized. This makes it harder to learn. True, but
my reflex to make things more complicated XXX

Ideally we would compute the difference between the predicted shape
and the expected shape, regardless of how they're drawn. Aside from
being a bit computationally challenging, this won't really work
because we need to attribute error directly to each output in order to
actually update the model in training. So I spent an afternoon writing
a routine to compute the best alignment of points between the actual
and expected outputs (Figure~\ref{fig:rotatej}). Aside from being
harder than it looked, my alignment code ended up being pretty slow
relative to the rest of training, even worse since it ran on the CPU
instead of GPU, and the training speed became UPSETTINGLY SLOW. So, I
discarded this excursion and threw this whole endeavour in the trash!
But I must have confused the Recycle~Bin icon with the fairly
complicated export-to-Truetype Font process that I built, because I
ran the model on the venerable Futura~\cite{futura} font and generated
{\bf Futurda} (Figure~\ref{fig:futurda}), which can be downloaded from my
website and installed for your desktop publishing needs.\footnote{Font
  downloads are available at \url{http://tom7.org/lowercase/}.}

% TODO: actually upload the fonts and make sure the URL is what you want

\begin{figure}[ht]
\begin{centering}
  \includegraphics[width=0.5 \linewidth]{rotatej}
\end{centering}
\caption{ Example alignment (from debug UI) of predicted shape (blue)
  to expected shape (green). We require each point to be mapped (red)
  to a point from the expected contour in a monotonic order (but
  several can be mapped to the same one), so that we can attribute
  error to each point. I hope it was fun to write this code because
  it was too slow to do pretty much anything but generate this figure!
} \label{fig:rotatej}
\end{figure}

% FIXME: Wait I had much better screenshots, see video/lowercase

% XXX add downloadable font too
\begin{figure*}[tp]
\begin{centering}
  \includegraphics[width=0.9 \linewidth]{futurda}
\end{centering}
\caption{ Type specimen for the generated font {\bf Futurda}. This is the
  classic font Futura, run through the vector based lowercasing model
  (Section~\ref{sec:vectorversion}). At top are Futura's letterforms
  \letterform{A}--\letterform{Z} run through the model and so ``made
  lowercase'' (it's obviously garbage). Next are
  \letterform{a}--\letterform{z},
  made even more lowercase. Also rubbish.
  At the bottom are the pangrams ``Dr.~Jock, TV Quiz Ph.D., bags few lynx''
  and ``Sphinx of black quartz, judge my vow!'' Although the text is
  illegible and kind of uninteresting, it does have a certain dynamic
  Rorschach aesthetic, like a collection of delicate moths pinned
  to paperboard, that one could consider framing or publishing in
  the proceedings of SIGBOVIK 2021.
} \label{fig:futurda}
\end{figure*}

@@

Don't give up! The fixed-size input/output of neural networks is
better suited to something like an array of pixels, and fonts can of
course be represented this way as well. To stay in the realm of what
my desktop computer with a single GeForce 1080 can do, I wanted to
keep the number of inputs and outputs pretty small. There's already an
excellent technique for representing font data as compact bitmaps,
which comes from computer graphics, called Signed Distance Fields
(SDFs)~\cite{green2007improved}. In a standard rasterization of a
font, each pixel of a bitmap contains 1 or 0, or perhaps an
anti-aliased value in-between. In an SDF, the bitmap instead contains
the distance to the nearest edge, and is signed (e.g. values inside
the shape are $> 0$, value outside are $< 0$). Actually in
practice we offset and saturate the values so that they are all in
$[0,1]$ (or bytes in $[0,255]$), with some nonzero ``on-edge value''
(say, 0.5) standing for ``distance 0''. In order to display the font
at the size of your choice, you then resize the SDF image with bilinear
interpolation, and then simply threshold the image. This works
surprisingly well (Figure~\ref{fig:sdf}).

\begin{figure}[ht]
  \includegraphics[width=0.95 \linewidth]{sdf-figure}
  \caption{
    The signed distance function representation of a letterform.
    At the very left, a $36\times 36$ pixel rasterization of
    the character without anti-aliasing, for comparison. Any
    scaling of this will have chunky pixel artifacts. Next,
    a $36\times 36$ pixel SDF of same. Third, simply scaling
    that $36\times 36$ image to $180 \times 180$ pixels with
    bilinear sampling. Finally, that image thresholded to
    produce a $180 \times 180$ pixel rasterization, which is
    far superior despite being derived from a $36 \times 36$ pixel
    bitmap. Typically this process is performed at an even higher
    scale and then downsampled to produce an anti-aliased image.
    } \label{fig:sdf}
\end{figure}

@@

I computed some stats on the font database, and determined the
following parameters for the fixed-size SDFs we train on. The images
are $36 \times 36$ pixels. The character box is placed such that there
are $2$ pixels of top padding, and $9$ pixels of left and bottom
padding. The character box is only ``nominal'' in the sense that the
font's contours can exceed its bounds, and this is completely normal
for a letter like \letterform{j} (which goes below the baseline and
often hangs to the left of the origin as well). I used an ``on-edge
value'' of 220 (because much more of the SDF is outside the letter
than inside) and the distance is scaled as 15 units per pixel (so that
pixels on the outer edge often have non-zero values). Compared to the
first version, I was somewhat more permissive in what fonts I trained
on, since there was no inherent limit to the number of contours their
complexity. I did exclude fonts whose rasterizations exceeded the
bounds of the SDF, which is possible (very wide \letterform{W} or
low-descending \letterform{j} perhaps) but rare.

@@

Having committed to the representation, again it is ``just'' a matter
of heating up the GPU to apply some linear and non-linear transforms.
The initial network had an input size of $36 \times 36 = 1296$ for the
SDF, and the output the same plus $26$ bonus outputs (one for each
letter, as before). I started with three hidden layers of $1296$, $1296$,
and $2916$ nodes, each sparse ($80\%$ of the weights are zero). Again,
don't take notes. This one works a bit better than before, but still
not impressive. The node references are assigned spatially (something
like the $20\%$ of the nodes on the previous layer that are closest to
the next layer's node) but due to a bug the spatial locality is
actually pretty strange.
% XXX if I show layer weights, forward-reference them here:
% It's responsible for the XXX
Every layer's transfer function is ``leaky relu'' ({\tt x > 0 \? x : 0.1 * x})
which is fast to compute and better than sigmoids in the output since
correct values will not just be 0 and 1. It would definitely make sense
to use convolutional layers for this problem, as features like serifs,
lines, curves, and so on could appear throughout the input and output.
I just haven't built support for that in my weird home-grown software,
yet.

>>>>

\usepackage{float}

% \usepackage{url}
% \usepackage{xurl}
\usepackage[pdftex]{hyperref}

% \usepackage{code}
% \usepackage{cite}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
% IPA symbols. safe turns off overrides for like \! which I still want
\usepackage[safe]{tipa}
\usepackage{chessboard}
% \usepackage{chessfs}
% \usepackage{adjustbox}

@@

\interfootnotelinepenalty=0
\setlength{\footnotesep}{2em}

\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}

% XXX the collection of crap above causes a first page
% with just "1" on it (probably conflicting packages). This
% magic just drops the first page, which is of course not
% the right solution, but...
\usepackage{atbegshi}% http://ctan.org/pkg/atbegshi
\AtBeginDocument{\AtBeginShipoutNext{\AtBeginShipoutDiscard}}

\newcommand\upsigbovik{
  \includegraphics[height=1em]{upsigbovik}
}

@@

\title{Lowestcase and Uppestcase letters: Advances in Derp Learning}
\author{Dr.~Tom~Murphy~VII~Ph.D.}\thanks{
Copyright \copyright\ 2021 the Regents of the Wikiplia Foundation.
Appears in \upsigbovik~2021 with the
OS2TypoLinegap of the Association for Computational Heresy; {\em IEEEEEE!}
press, Verlag-Verlag volume no.~0x40-2A. 1 em}

@@

If so, you made use of the fact that all letters have UPPERCASE
VERSIONS (e.g.~signifying ANGER) and lowercase versions
(e.g.~signifying u dont care lol). These dimensions have other uses,
for example, it is polite to start a person's name with a capital
letter to show that you took the time to regard their humanity (as it
takes extra work to press the caps lock key, press the first letter of
their name, and then press the caps lock key again to turn it off).
In German, Nouns start with uppercase Letters, signifying Superiority
over other grammatical Categories like Verbs and Adjectives. Lowercase
letters can be used to conserve printer ink. Actually, I'm not sure that
lowercase letters have any other uses, but let's just roll with it.

@@

% shift-reduce conflict
% Shift is often used to increase conflict, not reduce it

@@

We want to machine-learn~\cite{neuralnetwork} two functions,
\makelowercase\ and \makeuppercase. Each takes a letterform and
returns a letterform (we can choose how these are represented) and
does the needful, e.g.~\makelowercase(\letterform{A}) should return
\letterform{a}. In order to learn this function, we'll at least need a
lot of examples to use as training data. A training example for
\makelowercase\ is a letterform and its expected corresponding
lowercase one. We can ``easily'' find a large amount of examples by
using existing fonts, and pairing their \letterform{A} with their
\letterform{a}, and so on for all 26 letters, and symmetrically for
\makeuppercase.

@@

Next, my lord, do people have bad taste! And I say this as someone who
made dozens of amateurish fonts~\cite{dbzfonts} as a high school and
college student and who is contributing several new questionable
fonts\label{sec:newfonts} as a result of this paper. The database is
just filled with garbage that is unusable for this project: Fonts that
are completely illegible, fonts that are missing most of their
characters, fonts with millions of control points, Comic Sans MS,
fonts where every glyph is a drawing of a train, fonts where
everything is fine except that just the lowercase r has a width of
{\tt MAX\_INT}, and so on. So I built a UI
(Figure~\ref{fig:sortition}) for efficiently and mind-numbingly
cleaning up the database by marking fonts as broken or suitable (and
also categorizing them as serif, sans-serif, decorative, techno, etc.,
which classifications I never used). In doing this I noticed another
extremely common problem, which was that many fonts had the same
letter shapes for uppercase and lowercase letters. This would not do
for the current application!

\begin{figure}[tp]
\centering
  \includegraphics[width=0.98 \linewidth]{sortition}
\caption{ The interactive font data-cleaning UI. A seemingly endless
  series of fonts presents, with single keypresses putting the fonts
  into common categories such as (b)roken.
  % XXX maybe write more?
} \label{fig:sortition}
\end{figure}

@@

\paragraph{Black-box optimization.} If you have a function and want
to find arguments that minimize its output, the most efficient
techniques are generally those like gradient descent. (In fact, the
backpropagation algorithm we use to train the neural network in
Section~\ref{sec:neural} is gradient descent on the function that
takes the model weights and produces an error value for each output
node.) The problem with this is that you need to do some math to
compute the derivative of the function, and anyway you need to deal
with fiddly bits (Section~\ref{sec:fiddly}) unless the function is
convex and smooth, which it will not be. If you don't want to deal
with that, and have a fast computer (and who doesn't?), black-box
optimization algorithms are worth considering. Here, the
interface\footnote{ Here a simplfied wrapper around
  BiteOpt~\cite{biteopt} in my {\tt cc-lib} library. See
  \url{https://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/cc-lib/opt/}.}
is just something like (C++):

@@

\begin{figure}[tp]
\centering
  \includegraphics[width=0.9 \linewidth]{casealignment}
\caption{ Example alignment to reject the font {\tt
    DrippingGooExtended}. At left, \letterform{A} (red) and
  \letterform{a} (green) rendered with the identity transform, and
  their alignment ($35\%$ difference) below. At right, the transform
  found by the black-box optimizer and the resulting alignment with
  $1.7\%$ difference. Note that the shapes are still not an exact
  match (probably noise introduced in the font export process, which
  has to round the data to integers and might apply other non-linear
  transformations like curve simplification), but these are clearly
  not a useful pair for the current
  problem.} \label{fig:casealignment}
\end{figure}

@@

We'll at least put the data in a somewhat normalized form. The neural
network will take a fixed number of inputs to a fixed number of
outputs, so a simple approach is to decide on some maximum number
control points per letter, and only try training on fonts whose
letterforms fit within that budget. Letterforms can be made of
multiple contours (e.g.~a stacked \letterform{g} typically has two
holes in it, and \letterform{j} has two disjoint parts). I found that
most clean fonts had three or fewer contours, and when sorting them by
descending length, basically all of them fit within 100, 25, and 16
endpoints for the three. So, I only train on fonts where all of the
letters fit within this budget.\footnote{It would not be a good idea
  to reject only the letters that don't fit, because it might result
  in the network being trained on more \letterform{l}s (tends to be
  simple) than \letterform{g}s (tends to be complex).}

@@

Rather than try to work with both lines and B\'ezier curves, I
normalize each contour to only contain B\'eziers, by turning a line
segment into an equivalent B\'ezier with its control point at the
midpoint. This frees us from having to distinguish the two types in
the data. We also need each of the three contours to not be {\em too
  short}, so I fill out the fixed-size buffers by repeating the last
point. This is not great but does have the correct meaning (bunch of
useless zero-length edges). It has the property that any predicted
data can be rendered and has a straightforward point-by-point error
function (which might not be the case if we were predicting a dynamic
number of points).

The network I trained has an input layer size of $570 = (100 + 25 + 16)
\times 4 + 3 \times 2$ (one control point and one end point per
B\'ezier curve), plus a starting point for each of the three contours.
The output layer is the same size, plus 26 (see below). There are
three hidden layers of size 308, 308, 360. The first layer is dense
and the remainder are sparse,
%(two thirds to a half of the weights are zero),
for just about 1 million total parameters. All layers are leaky
rectified linear units ({\tt x > 0 ? x : 0.1 * x}), which is fast to
compute and better than sigmoids in the output since correct values
will not just be 0 and 1. If you're taking notes, don't, as again this
does not work well, and I don't know how people figure out what the
right network size is anyway. I just made it up. You can give {\em me}
your notes.

\paragraph{Bonus outputs.} The output includes the predicted shape,
and also 26 individual predictors for each of the 26 letters. So a
training example is actually like \letterform{C} $\rightarrow$
\letterform{c} $[0, 0, 1, 0, 0, \ldots, 0]$, with the 1 in the third
place because C is the third letter. We don't need these outputs for
the problem itself (e.g.~to lowercase new letter shapes), but there
are several ideas behind this. First, the lowercasing function we're
trying to learn does depend on the letter of the alphabet being
lowercased (in an extreme case, consider the lowercase-L
\letterform{l} and the uppercase-i \letterform{I}, which look the same
in many fonts but have different lowercase letterforms). By asking the
network to learn this as well (it is penalized when it gets the
prediction wrong), it must learn features that allow it to distinguish
different letters, and those features are available for use by outputs
we {\em do} care about. This is an easy way to coax it to learn
features that I know are meaningful without having to actually
engineer feature extractors by hand (or first train separate models,
etc.). Similarly, I could have asked it to predict whether the font
is italic, serif, the character's width, or anything else I have on
hand. Perhaps the most useful thing is that it's very clear what the
right answer is, so it gives me an easy way to see if the network is
learning anything {\em at all}. (It does.) Finally, we can do some
silly stuff with these; see Section~\ref{sec:hallucination}.

@@

\begin{figure}[ht]
\centering
  \includegraphics[width=0.9 \linewidth]{trainingvector}
\caption{ Screenshot of {\tt train.exe} running on the first vector-based
  version of the problem. Shown is the \makelowercase\ model's output
  (bottom) on four shapes from four fonts (top). Some dust in between
  is the activation of the network's layers. At the very bottom, the
  26 predictions for ``what letter is this?''. The output for
  \letterform{j} is not too bad; you can see the distinct dot (a
  separate contour) and it sort of looks like a \letterform{j}. The
  \letterform{e} also has two pieces as expected but is otherwise
  garbage. The model is unsure whether the second input is an H
  or a K, and has predicted a shape sort of ambiguously between those
  two. The \letterform{z} is also an embarrassment.
} \label{fig:trainingvector}
\end{figure}

I was so {\tiny confident} that this wouldn't work that I only trained
a \makelowercase\ model and didn't even worry about the more
complicated simultaneous training setup yet. I ran this for about
22,000 rounds, some 90 million training examples. Indeed it does not
work (Figure~\ref{fig:trainingvector}). It is not a total catastrophe,
though. We can tell from the 26 bonus outputs that the model can
clearly recognize letters (though perhaps just by memorization). Some
of the shapes it generates are along the right lines. ({\em Along the
  right lines}, get it??) I did not feel ANGRY at these results
because I expected it to not really work. Still, it ``has output'' and
so it can be used to generate a font. I made every glyph in Comic Sans
MS~\cite{comicsans} lowercase using the model (with the exception of the
\letterform{{\sf \%}} character, which has too many contours---{\it five}!).
Mostly this model produces small, non-confident scrawls, like little
grains of sand, so this font is called {\bf Comic Sands}
(Figure~\ref{fig:comicsands}). The TrueType version can be downloaded
from my website and installed for your desktop publishing
needs.\footnote{Font downloads are available at
  \url{http://tom7.org/lowercase/}.}

\begin{figure*}[tp]
\centering
  \includegraphics[width=0.95 \linewidth]{comicsands}
\caption{
  Type specimen for the generated font {\bf Comic Sands}. This is
  the hateful Comic Sans MS run through an early vector-based lowercasing
  model (Section~\ref{sec:vectorversion}). At top are Comic Sans's letterforms
  \letterform{A}--\letterform{Z} run through the model and so ``made
  lowercase'' (it's obviously garbage). Next are
  \letterform{a}--\letterform{z},
  made even more lowercase. Also rubbish.
  At the bottom are the illegible pangrams
  ``Dr.~Jock, TV Quiz Ph.D., bags few lynx''
  and ``Sphinx of black quartz, judge my vow!'' Although the output
  barely resembles letters, it does have a certain wispy
  Rorschach aesthetic, like a collection of delicate moths pinned
  to paperboard, that one could consider framing or publishing in
  the proceedings of SIGBOVIK 2021. It is certainly an improvement on
  the original font.
} \label{fig:comicsands}
\end{figure*}

\subsection{Just try making it more complicated!} \label{sec:complicated}

This problem of predicting the vector shape directly is a lot to ask
of a neural network, at least set up this way. One thing that did not
sit well with me is that the network could in principle generate a
perfect-looking result, but because it didn't have the points in the
expected order, it would be penalized. This makes it harder to learn,
and more prone to overfitting.\footnote{For example, imagine if the database
contains two versions of Helvetica that just have their points in a
different order---which is very likely the case btw---the model will
have to learn how to distinguish between these, but using information
we just don't care about.} This was one case where my questionable
reflex to make things more complicated did pay off!

First, I reduced the number of points in the input and output.
Reducing the dimension of the function being learned generally makes
learning a lot faster. This had the side-effect of reducing the number
of eligible fonts (by about half), and by nature these fonts are
simpler shapes. These effects alone could be responsible for the
improved performance of this second try.

I also output each contour's points in a normalized order, starting
from the point closest to the origin. This removes one needless degree
of freedom.\footnote{We can see (well, it's not pictured since I have
  far exceeded a reasonable number of figures in this paper, but {\em
    I} can see) how this manifests in the biases on the output layer,
  which are a proxy for the ``average prediction''. In the first
  model, because of the unstructured order, these are mostly near 0.5
  (center of the character) or 0.0 (degenerate, unused contours). In
  this new model, the distribution of biases is much more flat; it can
  learn that ``the first point tends to be near 0.25,0.25'' and ``the
  seventh point tends to be near 0.64,0.3.''}

Aside from the changes in the input (now 254 nodes) and output (280),
this second version has three sparse hidden layers of size 508, 508,
and 560 nodes; the first two are dense and the latter sparse. The
final model after some pruning had 609k parameters.

As this was training, I worked on another improvement. Ideally we
would compute the difference between the predicted shape and the
expected shape, regardless of how they're drawn. Aside from being a
bit computationally challenging, this won't really work because we
need to attribute error directly to each output in order to actually
update the model in training. I spent a valuable vacation day writing
a routine to compute the best alignment of points between the actual
and expected outputs (Figure~\ref{fig:trainingrotate}). Aside from being
harder than it looked, my alignment code ended up being pretty slow
relative to the rest of training, even worse since it ran on the CPU
instead of GPU, which reduced the training speed by $50\%$. I let it
run for 80,000 rounds, some 331 million training examples, but
eventually got bored of waiting on this approach that was slow to
train and seemed like a complicated version of an bad, oversimplified
approach. So, I control-C'd that thing and threw this whole endeavour
in the trash! But I must have confused the Recycle~Bin icon with the
fairly complicated export-to-TrueType Font process that I built,
because I ran the model on the venerable Futura~\cite{futura} font and
generated {\bf Futurda} (Figure~\ref{fig:futurda}).
%
% \begin{figure}[ht]
% \centering
%   \includegraphics[width=0.5 \linewidth]{rotatej}
% \caption{ Example alignment (from debug UI) of predicted shape (blue)
%   to expected shape (green). We require each point to be mapped (red)
%   to a point from the expected contour in a monotonic order (but
%   several can be mapped to the same one), so that we can attribute
%   error to each point. I hope it was fun to write this code because
%   it was too slow to do pretty much anything but generate this figure!
% } \label{fig:rotatej}
% \end{figure}
%

\begin{figure}[ht]
\centering
  \includegraphics[width=0.95 \linewidth]{trainingrotate}
\caption{ Screenshot (somewhat compacted) of training from near the
  final round of the vector model's training, illustrating the
  permissive loss function that finds the best alignment. At the bottom are
  the predicted lowercase shapes (blue), also shown with their
  expected shape (green). We require each point to be mapped (red)
  to a point from the expected contour in a monotonic order (but
  several can be mapped to the same one), so that we can attribute
  error to each point.
} \label{fig:trainingrotate}
\end{figure}

% XXX downloadable

\begin{figure*}[tp]
\centering
  \includegraphics[width=0.95 \linewidth]{futurda}
\caption{ Type specimen for the generated font {\bf Futurda}. This is
  the classic font Futura, run through the final, improved
  vector-based model (Section~\ref{sec:complicated}) to make each
  letter lowercase. The letterforms \letterform{A}--\letterform{Z}
  (top) become quite readable lowercase versions. The extra-lowercase
  \letterform{a}--\letterform{z} are also almost legible, but are
  mostly just scaled-down and screwed up versions of the lowercase
  letterforms. Could definitely imagine this appearing in the
  ``distressed fonts'' category of a 10001 Free TrueType Fonts CD-ROM
  in the 1990s, though.
} \label{fig:futurda}
\end{figure*}

@@

Don't give up! The fixed-size input/output of neural networks is
better suited to something like an array of pixels, and fonts can of
course be represented this way as well. To stay in the realm of what
my desktop computer with a single GeForce 1080 can do, I wanted to
keep the number of inputs and outputs pretty small. There's already an
excellent technique for representing font data as compact bitmaps,
which comes from computer graphics, called Signed Distance Fields
(SDFs)~\cite{green2007improved}. In a standard rasterization of a
font, each pixel of a bitmap contains 1 or 0, or perhaps an
anti-aliased value in-between. In an SDF, the bitmap instead contains
the distance to the nearest edge, and is signed (e.g.~values inside
the shape are $> 0$, value outside are $< 0$). Actually in
practice we offset and saturate the values so that they are all in
$[0,1]$ (or bytes in $[0,255]$), with some nonzero ``on-edge value''
(say, 0.5) standing for ``distance 0''. In order to display the font
at the size of your choice, you then resize the SDF image with bilinear
interpolation, and then simply threshold the image. This works
surprisingly well (Figure~\ref{fig:sdf}).

\begin{figure}[ht]
\centering
  \includegraphics[width=0.95 \linewidth]{sdf-figure}
\caption{
  The signed distance function representation of a letterform.
  At the very left, a $36\times 36$ pixel rasterization of
  the character without anti-aliasing, for comparison. Any
  scaling of this will have chunky pixel artifacts. Next,
  a $36\times 36$ pixel SDF of same. Third, simply scaling
  that $36\times 36$ image to $180 \times 180$ pixels with
  bilinear sampling. Finally, that image thresholded to
  produce a $180 \times 180$ pixel rasterization, which is
  far superior despite being derived from a $36 \times 36$ pixel
  image. Typically this process is performed at an even higher
  scale and then downsampled to produce an anti-aliased image.
} \label{fig:sdf}
\end{figure}

@@

I computed some stats on the font database, and determined the
following parameters for the fixed-size SDFs we train on. The images
are $36 \times 36$ pixels. The character box is placed such that there
are $2$ pixels of top padding, and $9$ pixels of left and bottom
padding. The character box is only ``nominal'' in the sense that the
font's contours can exceed its bounds, and this is completely normal
for a letter like \letterform{j} (which goes below the baseline and
often hangs to the left of the origin as well). I used an ``on-edge
value'' of 0.862 (because much more of the SDF is outside the letter
than inside) and the distance is scaled as 0.059 units per pixel
(chosen so that pixels on the outer edge often have non-zero values).
Compared to the first version, I was somewhat more permissive in what
fonts I trained on, since there was no inherent limit to the number of
contours or their complexity. I did exclude fonts whose rasterizations
exceeded the bounds of the SDF, which is possible (very wide
\letterform{W} or low-descending \letterform{j} perhaps) but rare.

@@

Having committed to the representation, again it is ``just'' a matter
of heating up the GPU to apply some linear and non-linear transforms.
The initial network had an input size of $36 \times 36 = 1296$ for the
SDF, and the output the same plus $26$ bonus outputs (one for each
letter, as before). I started with three hidden layers of $1296$, $1296$,
and $2916$ nodes, each sparse ($80\%$ of the weights are zero). Again,
don't take notes. This one works a bit better than before, but still
not impressive. The node references are assigned spatially (something
like the $20\%$ of the nodes on the previous layer that are closest to
the next layer's node) but due to a bug the spatial locality is
actually pretty strange.
% XXX if I show layer weights, forward-reference them here:
% It's responsible for the XXX
Every layer's transfer function is ``leaky relu'' again. It would
definitely make sense to use convolutional layers for this problem, as
features like serifs, lines, curves, and so on could appear throughout
the input and output. I just haven't built support for that in my
weird home-grown software, yet.
